---
title: "Soils of Lassen Volcanic National Park"
---

## Introduction 

Understanding the basic concepts of soil formation and genesis applied across landscapes is
difficult, it requires traversing the landscape and lots of digging or researching soil survey maps.
Finding, organizing, exploring and interpreting NRCS soil survey maps and data is a process
which consumes a lot of time and has evolved through the years from a strict physical format
into flexible digital formats. However, just because something is in a new digital format does
not always make the job easier as often times digital formats are stored in multiple online
databases and hosted with different web services.

The tools available from the NRCS include very large and robust yet clunky web soil survey app
(https://websoilsurvey.sc.egov.usda.gov/App/WebSoilSurvey.aspx), the soil data access portal
(https://sdmdataaccess.nrcs.usda.gov/) offers links to backend SQL services and web queryable
data from the data mart (https://ncsslabdatamart.sc.egov.usda.gov/) (Soil Survey, 2019). The
California resource soil lab at UC Davis has built a few higher level services to make this process
easier such as the SoilWeb apps (https://casoilresource.lawr.ucdavis.edu/soilweb-apps/)
(Beaudette et al., 2009). Additionally, climate and meteorological information in stored in
other databases which are accessed separately and must be joined with previous soil data
resources.

While the sheer access to raw data and information is enormous it takes a considerable amount
of time to organize, clean, and produce usable graphics from. There has been considerable
work in trying to simplify and improve workflow programmatically by the National Cooperative
Soil Survey. The algorithms for quantitative pedology package (soilAQP) for the R programming
language provides build in functions for querying, organizing, cleaning and producing tables
graphs from the multitudes of NRCS soil survey data available, as well as methods for querying
meteorological data (Beaudette et al., 2013) However, this requires an intermediate to
advanced knowledge of the R programming language as well as knowledge of querying
databases with SQL (R Core Team, 2013).

A simple and more direct method can be employed by referencing the paper copies of the soil
surveys produced by the NRCS, which are now also available in electronic PDF format and
contain most of the information required by a soil scientist to explore and learn about a new
and different landscapes in the United States.

If a curious soil scientist in potentia desired to learn more about the soils in the Cascade Range
of California north of the Sierra Nevada, Lassen Volcanic National Park seems like a good place
to start. Lassen Volcanic National Park is home to all 4 volcano types (cinder cone, plug dome,
shield, and stratovolcano). When coupled with other soil forming factors (climate, topography,
organics, time), the potential for an innovative and engaging opportunity for understanding
mechanisms of pedogenesis of soils in Lassen Volcanic National Park becomes clear especially
of less common soils such an Andisols. The objectives of this research is to
i) Learn about soil property relationships in Lassen Volcanic National Park Andisols
ii) Evaluate the soil survey data and information available and methods of use

## Methods

Soils in Lassen National Park were explored digitally through the use of SoilWeb to identify
potential trends and series to access data and gather more information about (Beaudette et al.,
2009). The soil Terracelake, Emeraldlake, Shadowlake, Juniperlake, and Buttelake series were
chosen as a subset of series to explore mostly to do the similarities of the “-lake” in their
naming structure. Four of the five soils are classified as Andisols, with Emeraldlake classified as
an Entisol with andic properties. The inclusion of a non Andisol soil may also serve as a subtle
contrast between the breakouts of an Andisol and an Entisol with Vitrandic properties (Table 5).
The soil datamart was accessed for soil series chemical data but did not have lab data for the
most of the soil series to be investigated. Soilweb provided access to soil taxonomy, physical
and some chemical data (horzons, particle size fractions, % organic matter, pH, CEC, and bulk
density) available (Table 5,6). The NRCS Soil Survey published in 2010 and was consulted to find
Andisol specific chemical information (NaF pH, %glass, P retention, and Al +1/2 Fe) and
appended to data found from the soil web database (Table 6)(Soil Survey, 2010). Soil chemical
and physical relationships such as were graphed in excel using the the data previously
organized (Figures 1-7).

Soil survey information from the CA078 survey area was queried from the NRCS Soil Data access
pathway with built in methods from the soilAQP package and used to create soil taxonomy and
map unit inclusion information in tables 1 – 4. Soil series profile sketches and dendrograms,
which show taxonomic relationships between series, were developed using the soilAQP
package as well (Figures 8 - 11) (Beaudette et al., 2013, Soil Survey, 2019).

The tables and graphs were analyzed for basic observable relationships and trends, statistical
methods were not employed to validate observations due to time constraints.

## Code 


```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)

## load tidyverse
library(tidyverse)

```

```{r}
## import lassen soils data csv
lassenSoils <- read_csv("LassenSoilData.csv")
```

```{r}
## CEC at pH 7 and soil pH

ggplot(data = lassenSoils) +
  geom_point(mapping = aes(x = ph, y = CEC, color=soilSeries))
```

```{r}
## CEC at pH 7 and percent clay

ggplot(data = lassenSoils) +
  geom_point(mapping = aes(x = clay_pct, y = CEC, color=soilSeries))
```

```{r}
## CEC at pH 7 and organic matter21 clay

ggplot(data = lassenSoils) +
  geom_point(mapping = aes(x = om_pct, y = CEC, color=soilSeries)) +
  coord_cartesian(xlim = c(0, 25))
```

```{r}
## CEC at pH 7 and percent clay

ggplot(data = lassenSoils) +
  geom_point(mapping = aes(x = om_pct, y = CEC, color=soilSeries))  +
  coord_cartesian(xlim = c(0, 25))

```

```{r}
## CEC change with Depth

ggplot(data = lassenSoils) +
  geom_point(mapping = aes(x = CEC, y = miid, color=soilSeries)) +
  scale_y_reverse()
```

```{r}
## load soilverse
library(aqp)
library(soilDB)
library(plyr)
library(sharpshootR)
library(knitr)
library(dplyr)
```



```{r}



```
### Soil AQP and Dendrograms
```{r}
## add functions to work with soil AQP data
# compute total component percentages for the current block of components
f.sum <- function(i) {
   n <- nrow(i) # number of rows
   s <- sum(i$comppct_r) # total component percent
   return(data.frame(pct=s, n=n)) # return the results
}

# pick the largest suborder from within each map unit
f.largest <- function(i) {
   i.sorted <- i[order(i$pct, decreasing=TRUE), ] # sort largest -> smallest
   top.suborder <- i.sorted$taxsuborder[1] # get the largest suborder
   top.suborder.pct <- i.sorted$pct[1] # get the the corresponding component percent
   return(data.frame(suborder=top.suborder, pct=top.suborder.pct)) # return results
}

```

```{r message=FALSE, warning=FALSE}
## Query data from soilweb API


# get map unit-level data
q.1 <- "SELECT mukey, muacres
FROM legend
INNER JOIN mapunit ON mapunit.lkey = legend.lkey
WHERE legend.areasymbol = 'CA789'"

# get component-level data
q.2 <- "SELECT 
component.mukey, cokey, comppct_r, compname, taxclname, taxorder, taxsuborder, taxgrtgroup, taxsubgrp
FROM legend
INNER JOIN mapunit ON mapunit.lkey = legend.lkey
LEFT OUTER JOIN component ON component.mukey = mapunit.mukey
WHERE legend.areasymbol = 'CA789'"

# submit queries
mu <- SDA_query(q.1)
co <- SDA_query(q.2)

# tabulate percentage of suborder-level taxa within each map unit (mukey)
l <- comp.suborder.sums <- ddply(co, .(mukey, taxsuborder), f.sum)

# keep the largest suborder, and its associated total percentage
comp.suborder <- ddply(l, .(mukey), f.largest)

# join the largest suborder (by map unit) to mu
x <- join(mu, comp.suborder, by='mukey')

# get an effective map unit area via: muacres * pct/100
x$effective.area <- with(x, muacres * pct / 100)


```

```{r cars}
sort(round(tapply(x$effective.area, x$suborder, sum)))

co %>% group_by(compname) %>% 
         dplyr::summarise(n=n()) %>% 
         kable()
```

```{r}

p <- sort(round(tapply(x$muacres, x$suborder, sum)))
p
```


```{r}

## attempt to make some summary stats for 

co %>% group_by(compname) %>% 
         dplyr::summarise( 
                   n=n(),
        n_d=n_distinct(taxorder)) %>% 
         kable()

co %>% group_by(taxorder) %>% 
         dplyr::summarise( 
                   n=n(),
        n_d=n_distinct(taxgrtgroup)) %>% 
         kable()


co %>% group_by(compname) %>% 
         dplyr::summarise( 
                   n=n()) %>% 
         kable()


co %>% filter(!is.na(taxgrtgroup)) %>% group_by(taxorder, taxsuborder, taxgrtgroup) %>% 
         dplyr::summarise( n_d=n_distinct(taxgrtgroup),
                   n=n()) %>% 
         kable()

co %>% group_by(taxorder, taxsuborder) %>% 
         dplyr::summarise(
                   n=n(),
                   n_d=n_distinct(taxgrtgroup)) %>%
    kable()
```

```{r}

## set to new variable:

t <- co %>% group_by(taxorder, taxsuborder) %>% 
         dplyr::summarise( 
                   n=n() )

 
 x %>% group_by(suborder) %>% 
         dplyr::summarise(sum=sum(muacres, na.rm = TRUE),
                   mean=mean(muacres, na.rm = TRUE),
                   n=n()) %>% 
         kable()
 
x %>% group_by(suborder) %>% 
         dplyr::summarise(sum=sum(muacres, na.rm = TRUE),
                   mean=mean(muacres, na.rm = TRUE),
                   n=n()) %>%
  kable()
```
```{r}
kable(addmargins(table(co$taxsuborder)))
```

```{r pressure}
sl.list <- c('terracelake', 'emeraldlake', 'shadowlake', 'buttelake', 'juniperlake')

# fetch and convert data into an SPC
sl <- fetchOSD(sl.list)

kssl.1 <- fetchKSSL('terracelake')
kssl.2 <- fetchKSSL('emeraldlake')
kssl.3 <- fetchKSSL('shadowlake')
kssl.4 <- fetchKSSL('buttelake')
kssl.5 <- fetchKSSL('juniperlake')
```

```{r}
g <- rbind(kssl.2, kssl.4, kssl.5)
table(g$taxmoistscl)

g$taxonname <- toupper(g$taxonname)

par(mar=c(1,1,4,3))
groupedProfilePlot(g, groups = 'taxorder', name='hzn_desgn',color='color', label='taxonname', id.style='side')


plot(sl, groups = 'taxorder', name='hzname', cex.names=0.65, axis.line.offset=-5)
plot(g, label = 'taxonname',name='hzn_desgn', color='moist_soil_color')

```


```{r}

# estimate soil depth based on horizon designation
sl$soil.depth <- profileApply(sl, estimateSoilDepth, name='hzname', top='top', bottom='bottom')

# plot profiles, order by soil depth
png("lassen_lake_order.png", width=1200, height=800, pointsize=33)
par(mar=c(0,0,0,0))
plot(sl, name='hzname', cex.names=0.65, axis.line.offset=-5)

# plot dendrogram + profiles
png("lassen_lake_dendrogram.png", width=1200, height=800, pointsize=33)
SoilTaxonomyDendrogram(sl, cex.taxon.labels=0.8)

```

```{r}
sx.list <- c('acroph', 'cragwash', 'buttewash', 'badgerflat', 'cenplat', 'dittmar', 'bearrubble', 'ashbutte','badgerwash', 'terracelake', 'emeraldlake', 'shadowlake', 'buttelake', 'juniperlake')

# fetch and convert data into an SPC
sx <- fetchOSD(sx.list)
SoilTaxonomyDendrogram(sx, cex.taxon.labels=0.8)
```

## Results and Discussion

The Lassen National Park Soil Survey covers 127,000 acres with 379 different map units which
includes 4 soil orders (Andisols, Entisols, Inceptisols, Histosols) composed of 8 sub orders which
break into 18 great groups (Table 1,2,3). Andisols make up around 85% of the soils in the park
(Table 2). Many of the map units are described at the series level of taxonomy, however, many
are only mapped as their great group class, which suggests the recent and raw nature of Lassen
and recent volcanic events (Table 4).

The relationship between soil pH and CEC was explored and did not express a strong
relationship, however a slight clustering of soils with low pH and high CEC can be seen in
Juniperlake and Shadowlake series (Figure 1). The weak association between soil pH and CEC
could be due to the interactions of soil clay content (% clay) and the influence of organic matter
(%OM) known to be acidic and were investigated for further information. Soil CEC and percent
clay content seem to exhibit a moderately strong positive relationship. CEC increased with
increasing clay content from a CEC of 0.4 cmolc/kg-soil at 0% clay content to an average CEC of
15 cmolc/kg-soil at around 4-6% clay and a maximum of 29 cmolc/kg at 2.8% clay content in the
Juniperlake series (Figure 2).

The relationship between %OM and CEC is very strong. CEC increased with %OM from 0.34
cmolc/mg-soil at 0.6% OM to a maximum of 29 cmolc./kg at 16% OM (Figure 3). Soil CEC
increased with soil %OM in a very linear fashion with the Juniperlake series at the high end with
most horizons above 20 cmolc/kg-soil and 8% OM. Organic matter typically builds up at the
surface and nd illuvial processes take time to draw organic matter and increase CEC at depth
which is reflected in the soil CEC as a function depth graph (Figure 4). CEC decreased with depth
in most profiles except the Buttelake series which increased dramatically in a lower 2Ab and
2Bw1,b Horizons which are buried horizons from recent volcanic events and are expressing the
same trend of decreasing CEC from the surface, except that it is a buried horizon (Figure 4).
Andic soil properties were investigated using NaF pH, % glass, and Al + ½Fe ammonium oxalate
extractable ion data. The relationship between NaF pH and soil CEC shows a weak to moderate
positive relationship with the soils with a NaF pH above 10 typically had a CEC above a 5
cmolc/kg-soil, which makes sense due to the NaF essentially exchanging with OH- ions bound
on exchange sites of short range order minerals (allophane, imogolite) typically found in
association with volcanic ashes.

The relationship between phosphorus retention and NaF is clear with a strong linear positive
relationship. The P retention in soils increased by 96% from a low of 6% retention and a NaF pH
around 8.2 to 100% P retention at NaF pH 10.5 – 11.5 (Figure 7). Phosphorus % retention with
NaF seems to indicate a deeper relationship with the short range order minerals (allophane,
imogolite) and has been a subject of interest with soil scientists studying Andisols.
The specifications of Andic soil properties as a diagnostic feature are explained in detail in the
NRCS Guide to Soil Taxonomy and one of the key components is the % glass to ammonium
oaxalate extractable Al + ½Fe fraction. Only 3 of the horizons from the “-lake” soil series
described fell outside of the dotted cutoff line (Figure 7). All of the soils meet the the
requirements for an andic diagnostic subsurface feature.

The soil profile sketches created using the soilAQP package are useful tools for interpreting soil
taxonomy and horizon information in a graphical way if you have a basic understanding of soil
horizon designations (Figure 7). When the the graphics are re-ordered as a dendrogram the
taxonomic relationships become clear and when coupled with horizon and color information a
good amount of information can be readily interpreted from a soil scientist in potentia (Figure
8). Soil profile sketches and dendrograms were developed to include another 9 soils from
Lassen Volcanic Park to further illustrate the taxonomic relationships of Andisols in Lassen
Volcanic National Park.

## Conclusions

A plethora of data exists for any soil scientist or soil enthusiast in a wide variety of formats.
Sometimes it is quicker to just read the manual soil survey and input data directly into excel, if
you know what you are looking for. Sometimes it is quicker and easier to download and
aggregate multiple soils, if you know how to do what you want to do. Ultimately it is up to the
user of the information to decide on how best to collect, organize, store, and synthesize new
data from the previous efforts of hundreds, if not thousands of people. This research employed
many different forms of data and used Microsoft Excel and R to interact and create useable
information for this report.

For the purposes of academic investigations and study utilizing or at least perusing all of the
resources available is useful. This investigation of Andisols was able to determine the relative
distribution of soil orders, sub orders, great groups and series within the park in a short amount
of time using programmatic methods, which would have much more difficult if tabulated and
calculated by hand (Tables 1-4). The Andic soil data that was not available on the soil datamart
regarding the series of interest and was input manually from reading the soil survey report
information in the series descriptions. The process took effort but yielded good results and nice
graphs which proved useful for interpretation (Tables 5,6; Figures 1-7).
The research into investigation and data collection methods of available soil survey data was a
valuable exercise which served to improved the knowledge of specific soil physical and chemical
relationships of andic soil properties. The research also highlighted the future work and goal to
improve understanding of the R programming language and the soilAQP package to be able to
readily develop actionable soil data in a short turnaround time.